\documentclass[10pt]{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Package Inclusion and Document Formatting %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage
{geometry,amsmath,amsthm,mathrsfs,amssymb,graphicx,bm,hyperref,url,pdfsync,
fancyhdr, wrapfig}
\pagestyle{fancy}
\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%
% Custom Commands %
%%%%%%%%%%%%%%%%%%%
\newcommand{\n}{\noindent}
\newcommand{\norm}[1]{\left\lvert#1\right\rvert}
\newcommand{\avg}[1]{\left\langle#1\right\rangle}
\newcommand{\abs}[1]{\left\vert#1\right\vert}
\newcommand{\figref}[1]{Figure \ref{#1}}

\newcommand{\HI}{H\,I\ }
\newcommand{\HII}{H\,II\ }
\newcommand{\Htwo}{H$_2$\ }
%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title Page Information %
%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Notes for PHYS 233: Interstellar Medium}
\author{Bill Wolf}
\date{\today}

\begin{document}

\vfill\maketitle\vfill \newpage

\tableofcontents \newpage

\section{Introduction}
\emph{Monday, January 6, 2013}\\

\n This course is called ``Interstellar Medium'' (ISM), but really it is about
the more general topic of the physics of diffuse gas in the universe. By
diffuse, we mean \emph{really} diffuse, down to number densities of 1
cm$^{-3}$, much more diffuse than the best vacuums in terrestrial laboratories.
This includes the interstellar gas in galaxies as well as the intergalactic gas
in between the galaxies.

\subsection{Organization of the ISM in the Milky Way} % (fold)
\label{sub:a_first_look}
Interstellar gas is what forms the stars in galaxies that are the dominant
sources of energy and light in the universe. Thus, understanding the physics of
the ISM helps us understand (and predict) the visible appearance of galaxies.\\

\noindent The space between the stars is occupied by gas, dust, cosmic rays,
electromagnetic radiation (from stars, the CMB, and radition from interstellar
matter), neutrinos, dark matter particles (whatever they are), and magnetic and
gravitational fields. Do note that dust is distinct from gas, as dust grains
have a lengthscale (micron) four orders of magnitude greater than gas particles
(angstrom).\\

\noindent In the Milky Way, there is roughly $10^{11}\ M_\odot$ in various
components:
\begin{itemize}
  \item $5\times 10^{10}\ M_\odot$ of stars
  \item $5\times 10^{10}\ M_\odot$ of dark matter
  \item $7\times 10^{9}\ M_\odot$ of interstellar gas
\end{itemize}
So the Milky Way is a rather gas-poor galaxy. Other galaxies like the
Magellanic clouds have most of their baryons in the gas phase. The obscuration
in the central disk of the Milky Way is produced by the dust along the midplane
(see color plate in 1.2, 1.65, 2.2 $\mu$m images). While the stars do form in a
very thin disk (thin compared to its width), the starlight heats up dust and
expands the dust disk to higher altitudes from the midplane.\\

\n The magnetic fields in the Milky Way can be detected by looking for
synchrotron emission (caused by electrons spiraling around galactic field
lines). We do observe this emission (see additional plate) and believe the
magnetic fields are caused by a dynamo effect due to the circulating galactic
material.

\subsubsection{Gas Content} % (fold)
\label{ssub:gas_content}
There is about $5\times 10^9\ M_\odot$ of hydrogen in the Milky Way, in the
following states
\begin{itemize}
  \item $2.9\times 10^9\ M_\odot$ of \HI gas (neutral)
  \item $1.12\times 10^9\ M_\odot$ of \HII gas (ionized)
  \item $0.84\times 10^9\ M_\odot$ of \Htwo gas (molecular)
\end{itemize}
We can track the presence of interstellar gas by looking at C$^+$ emission in
the Milky Way, which tracks hydrogen presence very well, though the weighting
is different since the emission strength is proportional to the density
squared.\\

\n Finally, we track galactic gas via CO emission, which is very concentrated
to the disk, but also has strong filaments outside the central disk of the
galaxy.\\

\n The interestellar medium exhibits turbulence that is the reult of supernovae
and other violent hydrodynamic events.
% subsubsection gas_content (end)

\paragraph{Phases of Interstellar Gas} % (fold)
\label{ssub:phases_of_interstellar_gas}
We categorize the ISM into various phases according to the ionization/molecular
state of its hydrogen (descriptions taken from Draine 1.1):
  \begin{itemize}
    \item \textbf{Warm \HI}: Neutral hydrogen gas at temperatures around $T
    \approx 10^{3.7}\ \mathrm{K}$. Typically at densities of
    $n_{\mathrm{H}}\approx 0.6\ \mathrm{cm^{-3}}$. Occupies about 40\% of the
    galactic disk. Often referred to as the warm neutral medium (WNM)
    \item \textbf{Cool \HI}: Neutral hydrogen gas at temperatures around $T
    \approx 100$ K with higher densities of $n_{\mathrm{H}}\approx 30\
    \mathrm{cm^{-3}}$. Occupies around 1\% of the local ISM, sometimes referred
    to as the cold neutral medium (CNM)
    \item \textbf{Diffuse H}$_{2}$: Like the CNM, but with larger densities and
    column densities, allowing molecular hydrogen (\Htwo) to be more abundant.
    \item \textbf{Dense H}$_{2}$: Gravitationally bound clouds with number
    densities exceeding $10^3\ \mathrm{cm^{-3}}$. Distinguished by their dark
    appearance (strong optical extinction) in their central regions. Hosts of
    star formation. Not ``dense'' by terrestrial standards.
    \item \textbf{Ionized \HII} at $10^4$ K: Gas consisting largely of
    photoionized hydrogen, likely from ultraviolet radiation from a nearby
    massive star. May be dense material from a cloud (\HII region) or diffuse
    intercloud material (diffuse \HII). Lifetimes typically on the order of
    that of a massive star (Myrs). Extended diffuse photoionized regions,
    sometimes called the warm ionized medium (WIM), contain \emph{much} more
    total mass than the more visible \HII regions. Planetary nebulae (PNe) are
    also in this class.
    \item \textbf{Coronal Gas} (Ionized \HII at $\log T > 5.5$): Very hot,
    ionized gas that has been shock-heated by blastwaves from supernova
    explosions. It is mostly collisionally ionized with some exotic ionization
    states present. Typically very low density, filling about half the volume
    of the galactic disk. Cools on Myr timescales. Often referred to as the hot
    ionized medium (HIM).
  \end{itemize}
% paragraph phases_of_interstellar_gas (end)
\paragraph{Elemental Composition} % (fold)
\label{par:elemental_composition}
The gas is largely made of hydrogen and helium from the early universe, but an
additional 1-2\% (by mass) is heavy elements ($Z>2$), or to astronomers,
``metals''. The metallicity is actually a declining function from the center of
the galaxy outwards since it is the result of stellar processing, and the
center of the galaxy is more evolved then its relatively young outer regions.
The metallicity near the solar system is about half of that at the galactic
center. We know the local composition of the ISM by measuring solar
photospheric abundances and the composition of meteorites (See table 1.4 in
Draine).
% paragraph elemental_composition (end)

\subsubsection{Energy Content} % (fold)
\label{ssub:energy_content}
Energy in the CMB has a strange equipartition where nearly all components are
near 1 eV cm$^{-3}$. Some of this is self-sustaining, but since the CMB energy
density increases as $(1+z)^4$, its equality is coincidental. The rough energy
densities by categories are given in Table~\ref{tab:1.1}. Since the galactic
magnetic fields are driven by bulk motions, it is unsurprising that the
turbulent (bulk) and magnetic energies are nearly equal. If the cosmic ray
energy density were significantly greater, the magnetic fields could not
confine them, so they would simply escape the galaxy. Similarly, if the energy
from starlight were significantly higher, the ISM would expand and likely
dampen star formation, causing the near equipartion that we observe.
\begin{table}
  \centering
  \begin{tabular}{l l}
    Energy Type & Density (eV $\mathrm{cm^{-3}}$)\\
    \hline
    Thermal & 0.49\\
    Bulk Kinetic & 0.22\\
    Cosmic Ray & 1.39\\
    Magnetic & 0.89\\
    CMB & 0.265\\
    Far infrared from dust & 0.31 \\
    Starlight & 0.54
  \end{tabular}
  \caption{Energy densities in the ISM.}
  \label{tab:1.1}
\end{table}
If any of these components grows much larger than the gravitational binding
energy, then hydrostatic equilibrium is disrupted, driving a wind. All galaxies
drive winds at some time.
% subsubsection energy_content (end)
\subsubsection{Galactic Endgame} % (fold)
\label{ssub:galactic_endgame}
All of the ISM constituents are present between galaxies, and the same physical
processes apply to studying the intergalactic medium (IGM). In our present
state of the field, we can't account for the relative overabundance of dark
matter in the vicinity of a galaxy. It is thought that the ``missing baryons''
are in the circum-galactic material (CGM). The only place where the mass is
``right'' is in massive clusters. What is unknown is what sort of lengthscale
determines how far out this CGM might extend. Interestingly, the material
observed in the CGM contains metals, so either galactic winds are responsible
for pushing material processed by stellar evolution out of the galaxy, or very
massive population III stars formed at $z\sim 20-40$ to form the first metals.
% subsubsection galactic_endgame (end)
% subsection a_first_look (end)
\subsection{Basics of Diffuse Gas} % (fold)
\label{sub:basics_of_diffuse_gas}
Perhaps the prettiest regions of diffuse gas are the \HII regions, like the
Orion Nebula, which is illuminated by four central stars (the trapezium). The
typical \HII region has a number density around $1\ \mathrm{cm^{-3}}$ and a
temperature of $10^4$\ K. A big difference between these \HII regions (and in
fact most instances of interstellar gas) and stellar interiors is that they are
\emph{not} in local thermodynamic equilibrium (LTE).\\

\n The NLTE-ness of diffuse gas iss due tot he much, much lower collision rate
than in planetary atmospheres or stellar interiors. In order to achieve LTE, a
gas must satisfy the following conditions:
\begin{itemize}
  \item Population of excited states given by Boltzmann equilibrium
  \item Particle energies distributed according to the Maxwell distribution
  \item Ionization balance given by Saha Equation
  \item Photon energies described by Planck Function
\end{itemize}
In diffuse gas, we may indeed be able to identify a kinetic temperature that
gives the average kinetic energy of the particles, but this temperature need
not be the same as the Boltzmann temperature, which gives the occupation of
energy states in the system. Likewise, the ionization temperature, Planck
temperature (of a blackbody), and other such derived temperatures need not
match such a diffuse system.
% subsection basics_of_diffuse_gas (end)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Wednesday, January 8, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Collisionally Excited Lines} % (fold)
\label{sec:collisionally_excited_lines} % (fold)
\n\textit{Wednesday, January 8, 2014}\\

\begin{table}
  \centering
  \begin{tabular}{l l l}
  Environment & Number Density (cm$^{-3}$) & Collision Timescale (s)\\
  \hline
  Earth Atmosphere & $2\times 10^{13}$ & $\sim 10^{-9}$\\
  Molecular Cloud & $10^6$ & $10^8$ (several days)\\
  ISM & 1 & $10^{11}$ ($10^4$ yr)
  \end{tabular}
  \caption{Some characteristic collision timescales.}
  \label{tab:2.1}
\end{table}
\n A typical timescale for collisions between particles can be estimated via
\begin{equation}
  \label{eq:coll:1} \tau = \frac{1}{\sigma v n}
\end{equation}
That is, it's essentially the inverse of the expected collision rate. Here
$\sigma$ is the cross section for interaction, $v$ is the relative velocity of
particles, and $n$ is the number density of particles. Some typical timescale
for various environments are given in Table~\ref{tab:2.1}.\\

\n Also of importance is the volumetric rate of emission or photons due to
recombination, which will be proportional to the product of electron and ion
number densities, or $r_{\mathrm{emis}}\propto n_{\mathrm{ion}}n_e$. We define
the \textbf{Emission Measure} as
\begin{equation}
  \label{eq:coll:2} \mathrm{EM} = \int n_e^2 ds
\end{equation}
where the integral is over some path in space. The ``natural'' units for this
measure is then pc cm$^{-6}$.
\paragraph{Example: Emission Measure of a Nova Shell} % (fold)
\label{par:example_emission_measure_of_a_nova_shell}
  For a nova shell with mass $\Delta M = 10^{-4}\ M_\odot$ expanding with a
  velocity $v=10^3\ \mathrm{km\ s^{-1}}$, we get an approximate emission
  measure of
  \begin{equation}
    \mathrm{EM} = (3\times 10^{-4}\ \mathrm{pc})(10^7\ \mathrm{cm^{-3}}) \sim
    3\times 10^{10}\ \mathrm{pc\ cm^{-6}}
  \end{equation}
  So at early times nova shells are very bright. We could play this game with
  other objects like planetary nebulae (PNe), \HII Regions, Diffuse warm-ionized
  gas, or circumgalactic gas to get an idea of how bright these objects would
  appear.
% paragraph example_emission_measure_of_a_nova_shell (end)
\subsection{Two-Level Atom} % (fold)
\label{sub:two_level_atom}
  % TODO Add energy level diagram
  If we consider an atom that has only two states, 1 and 2, with degeneracies
  $g_1$ and $g_2$ with an energy gap $\epsilon_{12}$, the collisional cross
  section is given by
  \begin{equation}
    \label{eq:coll:3} \sigma_{12}(E) = \frac{h^2}{8\pi m_e
    E}\frac{\Omega_{12}(E)}{g_1}
  \end{equation}
  where $\Omega$ is the \textbf{collision strength}, which is a dimensionless
  quantity with values around order unity. The \textbf{Principle of Detailed
  Balance}, which we have not yet proven, tells us that
  $\Omega_{12}=\Omega_{21}$.\\
  
  \n Indeed, the principle of detailed balance tells us that the rate of upwards and downward collisional transitions are equal, or
  \begin{align}
    \label{eq:coll:4}R_{12} &= R_{21}\\
    \label{eq:coll:5}n_1 n_e \alpha_{12} &= n_2 n_e \alpha_{21}
  \end{align}
  where $\alpha$ is the collisional excitation coefficient for that reaction. To derive this principle, we turn to statistical mechanics, and the concept of a \textbf{partition function}, which is the normalization to a boltzmann distribution:
  \begin{equation}
    \label{eq:coll:6} Z(T) \equiv \sum_s e^{-E(s)/(kT)}
  \end{equation}
  % TODO fix this mess
  where the sum is over all microstates. In a diulte gas, the partition function is actually a product of the particion functions from the internal system and that of the translational system:
  \begin{equation}
    \label{eq:coll:7} Z(T) = Z_{\mathrm{trans}}(T) \times Z_{\mathrm{int}}(T)
  \end{equation}
  Then we can define the particion function per unit volume in the usual way
  \begin{align}
    \label{eq:coll:8} f(T) &= \frac{Z(T)}{V}\\
    \label{eq:coll:9} &= \left(\frac{2\pi M_x kT}{h^3}\right)^{3/2}\times Z_{\mathrm{int}}(T)
  \end{align}
  Where we've glossed over some details present in Draine (Chapter 3). Using statistical mechanics, we are more or less assuming we are in LTE, so the number densities of various states for the reaction
  \begin{equation}
    \label{eq:coll:10} A + B \rightleftharpoons C
  \end{equation}
  is given in the usual way:
  \begin{equation}
    \label{eq:coll:11} \frac{n_{\mathrm{LTE}}(C)}{n_{\mathrm{LTE}}(A)
    n_{\mathrm{LTE}}(B)} = \frac{f(C)}{f(A)f(B)}
  \end{equation}
  which constrains the rate coefficients in a specific way. 
  % TODO something about how we don't need LTE
  Armed with this tool, we can actually try to compute collisional excitation rates:
  \begin{align}
    \label{eq:coll:12} R_{12} &= n_1 n_e \int_{E_{\mathrm{12}}}^\infty \sigma(v)v\,f(v)\,dv\\
    \label{eq:coll:13} R_{21} &= n_2 n_e \int_0^\infty \sigma(v)v\,f(v)\,dv
  \end{align}
  Now using the principle of detailed balance, we equate these rates and get 
  \begin{equation}
    \label{eq:coll:14} n_1e^{-E_{12}/(kT)}\frac{\Omega_{12}}{g_1} = \frac{n_2 \Omega_{21}}{g_2}
  \end{equation}
  Solving for the ratio of the number densities, we get
  \begin{equation}
    \label{eq:coll:15} \frac{n_2}{n_1} = \frac{\alpha_{12}}{\alpha_{21}} = \frac{\Omega_{21}}{g_1}\frac{g_2}{\Omega_{12}} e^{-E_{12}/(kT)}
  \end{equation}
  Or, more simply
  \begin{equation}
    \label{eq:coll:16} \frac{n_2}{n_1} = \frac{\Omega_{21}}{\Omega_{12}} \frac{g_2}{g_1} e^{-E_{12}/(kT)}
  \end{equation}
  Now if we are in LTE, we immediately recover $\Omega_{21} = \Omega_{12}$ since the Boltzmann ratio is already present. However, the collision strength is an intrinsic property of the ion, so this must be true in all cases. That is, we dont' require LTE to recover $\Omega_{12}=\Omega_{21}$ simply because we proved that it \emph{does} apply in LTE, and thus at all times.\\
  
\n In the low density limit, spontaneous decay dominates emission since collisionally excited transitions are few and far between. Then the transition rate can be written as
\begin{equation}
  \label{eq:einstein:1} n_en_1\alpha_{12} = n_2 A_{21}
\end{equation}
where $A$ is the \textbf{Einstein Coefficient} for spontaneous decay from state 2 to state 1. 
% TODO explain this next equation
\begin{equation}
  \label{eq:einstein:2} F_{12} = E_{12} A_{21} n_2
\end{equation}

Subbing in some Boltzmann algebra, $n_2$ can be expressed as
\begin{equation}
  \label{eq:einstein:3} n_2 = \frac{n_en_1}{A_{21}} \left( \frac{2\pi \hbar^4} {k m_e^3}\right)^{1/2} \frac{\Omega_{12}}{g_1} \frac{e^{-E_12}/(kT)}{\sqrt{T}}
\end{equation}
Plugging \eqref{eq:einstein:3} into \eqref{eq:einstein:2} and evaluating some of the constatns, we get
\begin{equation}
  \label{eq:einstein:4} F_{12} = E_{21} n_e n_1 \left(8.62942\times 10^{-6}\right) \frac{1}{\sqrt{T}} \frac{\Omega_{12}}{g_1}e^{-E_{12}/(kT)}
\end{equation}
Now in the high density limit, we do something similar, getting
\begin{equation}
  \label{eq:einstein:5} n_e n_1 \alpha_{12} = n_2 A_{21} + n_e n_2 \alpha{21}
\end{equation}
Now we get
\begin{equation}
  \label{eq:einstein:6} F_{12} = E_{21} A_{21} n_1 \frac{g_2}{g_1} e^{-E_{12} / (kT)}
\end{equation}
These two regimes beg for us to define a critical density that separates these two limits where $n_2A_{21} = n_2n_e \alpha_{21}$, which is, after some algebra,
\begin{equation}
  \label{eq:einstein:7} \boxed{n_{e, \mathrm{crit}} = \frac{A_{21}g_2 T^{1/2}} {\beta \Omega_{12}}}
\end{equation}
where 
\begin{equation}
  \label{eq:einstein:8} \beta \equiv \left[\frac{2\pi \hbar^4}{k m_e^3}\right]^{1/2}
\end{equation}
For forbidden lines, this critical density is in the range of $10^2-10^7\ \mathrm{cm^{-3}}$. For intercombination lines, the critical density is around $10^10\ \mathrm{cm^{-3}}$. For resonance lines (aka permitted lines), the critical density is $10^{15}\ \mathrm{cm^{-3}}$
% subsection two_level_atom (end)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Friday, January 10, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Three-Level Atom} % (fold)
\label{sub:three_level_atom}
\textit{Friday, January 10, 2014}
\subsubsection*{Suggested References} % (fold)
\label{ssub:suggested_references}
\begin{itemize}
  \item Dopita \& Sutherland Chapter 3
\end{itemize}
% subsubsection suggested_references (end)
Now we consider the slightly more complicated three-level atom. Considering that the collisional excitation rate is, in general, given by
\begin{equation}
  \label{eq:three:1} C_{\mathrm{ij}} = n_e \alpha_{ij}
\end{equation}
for the appropriate rate factors $\alpha_{ij}$ for excitations from state $i$ to state $j$. Now we'll employ statistical equilibrium to figure out the equilibrium rates and concentrations. For level 3, we get the following
\begin{equation}
  \label{eq:three:2} n_1C_{13} + n_2 C_{23} = n_3(C_{31} + C_{32} + A_{32} + A_{31})
\end{equation}
now for level 2, we get
\begin{equation}
  \label{eq:three:3} n_1 C_{12} + n_3(C_{32} + A_{32}) = n_2(C_{21} + C_{23} + A_{21})
\end{equation}
and for normalization purposes, let's assume that the total number density is given by $n$:
\begin{equation}
  \label{eq:three:4} n_1 + n_2 + n_3 = n
\end{equation}
\eqref{eq:three:2} and \eqref{eq:three:3} essentially just equate the rates of reactions (collisional or spontaneous) for the ``formation'' of that energy state (left side) to the rates of ``destruction'' rates (right side).\\

\n Now we'll make some assumptions, namely that the energy levels are roughly equally spaced, or $E_{12} \approx E_{23}$. This tells us that the rate of excitations from state 1 to state 3 will be much lower than that from state 1 to state 2, or $C_{13}\gg C_{12}$.
% TODO Why do we make this assumption?
Now in a low density limit, we may ignore collisional excitation, which reduces our equations to a much cleaner form:
\begin{align}
  \label{eq:three:5} n_1 C_{13} &= n_3(A_{32}+A_{31})\\
  \label{eq:three:6} n_1 C_{12} + n_3A_{32} &= n_2A_{21}\\
  \label{eq:three:7} n_1 + n_2 + n_3 &= n
\end{align}
Solving these equations for $n_3$ and $n_2$ gives
\begin{align}
  \label{eq:three:9} n_3 &= \frac{n_1 C_{13}}{A_{32} + A_{31}}\\
  \label{eq:three:8} n_2 &= \left[n_1 C_{12} + \frac{A_{32}}{A_{32} + A_{31}} n_1 C_{13}\right]\frac{1}{A_{21}}\\
  \label{eq:three:10} &\approx \frac{n_1 C_{12}}{A_{21}}
\end{align}
Then the ratio of fluxes is 
\begin{equation}
  \label{eq:three:11} \frac{F_{32}}{F_{21}} = \frac{n_3 A_{32} E_{32}}{n_2 A_{21}E_{21}} = \cdots = \frac{C_{13}}{C_{12}} \frac{A_{32}}{A_{32} + A_{31}} \frac{E_{32}}{E_{21}}
\end{equation}
Now recalling that the $C$'s are related to our $\alpha$'s, via, for example,
\begin{equation}
  \label{eq:three:12} C_{13} = n_3 \left(\frac{2\pi \hbar^4}{k m_e^3}\right)^{1/2} \frac{1}{\sqrt{T}} \frac{\Omega_{13}}{g_1} e^{-{E_{13}}/kT},
\end{equation}
\eqref{eq:three:11} reduces to a function of a bunch of physical constants and the temperature
\begin{equation}
  \label{eq:three:13} \boxed{\frac{F_{32}}{F_{21}} = \frac{E_{32}}{E_{21}} \frac{A_{32}}{A_{32}+A_{21}} \frac{\Omega_{13}}{\Omega_{12}} e^{-E_{23}/kT}}
\end{equation}
So now given a line strength ratio, we can actually compute a temperature for that diffuse gas. Some examples of ions where these assumptions are valid are N$^+$, O$^{+2}$, Ne$^{+4}$, and S$^{+2}$.\\

\n Now instead if we assume that $E_{23} \ll E_{12}$ (i.e., the second and third states are very close in energy). This means, though it's not obvious, that $A_{32} \ll A_{31}$. Still in the low density limit (ignoring collisional de-excitation), we'll simplify \eqref{eq:three:2} and \eqref{eq:three:3} again:
\begin{align}
  \label{eq:three:14} n_1 C_{13} &= n_3(A_{32} + A_{31})\\
  \label{eq:three:15} & \approx n_3A_{31}\\
  \label{eq:three:16} n_1 C_{12} &= n_2 A_{21}
\end{align}
Then the flux ratios are simply
\begin{equation}
  \label{eq:three:17} \frac{F_{31}}{F_{21}} = \frac{n_1 C_{13}}{A_{31}} \frac{A_{21}}{n_1 C_{12}} \frac{E_{31}}{E_{21}} \frac{A_{31}}{A_{21}} \approx \frac{C_{13}}{C_{12}}
\end{equation}
So the flux rates depend \emph{only} on the collisional excitation rates in this limit. Re-expressing the flux ratio in terms of more fundamental quantities:
\begin{equation}
  \label{eq:three:18} \frac{F_{31}}{F_{21}} \approx \frac{C_{13}}{C_{12}} \approx \frac{\Omega_{13} e^{-E_{13}/kT}}{\Omega_{12} e^{-E_{12}/kT}} \approx \frac{\Omega_{13}}{\Omega_{12}} \approx \frac{g_{3}}{g_2}
\end{equation}
for somewhat unobvious reasons. So if we measure line strengths and get this magic number, all we really know is that we're in the low density limit. The magic comes in when we are in the high density limit:
\begin{equation}
  \label{eq:three:19} \frac{F_{31}}{F_{21}} = \frac{n_3}{n_2} \frac{A_{31}}{A_{21}} \frac{E_{31}}{E_{21}}
\end{equation}
Since in the high density limit there are many collisions, the number densities should be Boltzmann-like:
\begin{equation}
  \label{eq:three:20} \frac{n_3}{n_2} = \frac{g_3}{g_2} e^{-E_{23}/kT}\approx \frac{g_3}{g_2}
\end{equation}
So the flux ratio is now
\begin{equation}
  \label{eq:three:21} \frac{F_{31}}{F_{21}} \approx \frac{A_{31}}{A_{21}}\frac{g_3}{g_2}
\end{equation}
So far, this isn't really a density diagnostic, though we've identified the values at both limits. The transition between these limits should be at the critical density discusse earlier, defined by:
\begin{equation}
  \label{eq:three:22} A_{21} n_2 = n_e n_2 \alpha_{21}
\end{equation}
Earlier, we found that this critical density was found to be
\begin{equation}
  \label{eq:three:23} n_{e, \mathrm{crit}} = \frac{A_{21}g_2 T^{1/2}}{\beta \Omega_{12}}
\end{equation}
Different transitions will have different critical densities that depend on their intrinsic physics, for electron densities below the smaller critical density, the flux ratio is constant, and the same holds for electron densities greater than the larger critical density. In between, though, their is a transition zone wherein we can approximate a density. Examples of ions that allow for such a diagnostic are O$^+$, S$^+$, Ne$^{+3}$, and Ar$^{+3}$.
% subsection three_level_atom (end)
\subsection{Energy Level Diagrams} % (fold)
\subsubsection*{Suggested References} % (fold)
\label{ssub:suggested_references}
\begin{itemize}
  \item Draine Chapter 4
\end{itemize}
% subsubsection suggested_references (end)
\label{sub:energy_level_diagrams}
Often we will refer to ions by their atomic name followed by a roman numeral
that is one greater than its oxidation state. For instance, O$^{+2}$ is
referred to as O\,III.
% TODO Add energy level diagrams
There will be some energy level diagram figures after this demonstrating where
some of these energy level assumptions work.\\

\n Let's briefly review the electronic configurations of some of these ions.
O\,I has 8 electrons, so in its ground state it has an electronic configuration
of $1s^2\ 2s^2\ 2p^4$. O\,III however, has a configuration of $1s^2\ 2s^2\
2p^2$, where the outermost electrons are actually in different orbitals
(remember Hund's rules!). This $p^2$ configuration disallows the two outermost
electrons from having the same quantum numbers ($n,\ \ell,\ m,\ s$). In this
case, we know that $n=2$ (the ``2'' in $2p^2$), and $\ell=\pm1$ (the $p$). For
electrons, $s = \pm 1/2$. So the total spin can take on values of 0, or 1 and
the total orbital angular momemtum can be 0, 1, or 2. We'll often see this
configurations labeled via $\mathrm{^{2S+1}L_J}$ where S is the total spin
angular momentum, L is the orbital name ($s,\ p,\ d,\ f,$ etc.), and J is the
total angular momentum. The available states for O\,II are shown in
Table~\ref{tab:oii}
\begin{table}
  \centering
  \begin{tabular}{l l l}
    $^1\mathrm{D}_{J=2}$ & L = 2 & S = 0\\
    $^3\mathrm{D}_{J=3,2,1}$ & L = 2 & S = 1 \\
    $^1\mathrm{P}_{J=1}$ & L = 1 & S = 0\\
    $^3\mathrm{P}_{J=2,1,0}$ & L = 1 & S = 1\\
    $^1\mathrm{S}_{J=0}$ & L = 0 & S = 0\\
    $^3\mathrm{S}_{J=1}$ & L = 0 & S = 1
  \end{tabular}
  \caption{Possible electron configurations for O\,II.}
  \label{tab:oii}
\end{table}
  Not all of these states are allowed transitions though, for reasons I still
  need to record here.
  % TODO Explain transition rules.

% subsection energy_level_diagrams (end)
\subsection{Infrared Line Diagnostics} % (fold)
\label{sub:infrared_line_diagnostics}
Transitions between fine-structure levels of $p^2$ and $p^4$ ions are the
dominant cooling processes in gas at $T = 100-3000$ K. Examples of such lines
are [CII] 158 $\mu$m and [OIII] 88.36 $\mu$m. Atmospheric water vapor blocks
these lines in the 25 to 300 $\mu$m range, but airborn observatories can
observe them, and ALMA is making great headway in observing mm-scale
radiation.\\
% TODO fill this out... missed information from the slides.

\n However, note that in the real world (like in \HII regions), the ISM is
clumpy, so these diagnostics actually only measure the number density of the
clumps, $n_{e,c}$. The emission measure, however, cares about the average
electron density: $\mathrm{EM}\sim \avg{n_e^2}\times(\mathrm{length})$. We
parameterize this clumpiness with a \textbf{volume filling factor} for the
clumps, $f$, defined by
\begin{equation}
  \label{eq:ff:1} \avg{n_e^2}\approx fn_{e,c}^2
\end{equation}
Typically we find $f\sim 0.1-0.1$ in nearby \HII regions.
% subsection infrared_line_diagnostics (end)
\subsection{Resonance Lines and Selection Rules} % (fold)
\label{sub:resonance_lines_and_selection_rules}
Remember that energy transitions are surrounded in square brackets when they
are the so-called ``forbidden transitions''. The more run-of-the-mill resonance
transitions, or allowed transitions, are those that obey the following rules:
\begin{enumerate}
  \item Only 1 electron involved in the transition
  \item Initial and final states have different parity
  \item Emitted photon carries 1 unit of angular momentum, so $\Delta L = \pm 1$
  \item Electron spin does not change
  \item Change in the total angular momentum of the active electron is $\Delta
  J = \pm 1$ or $0$, with $J=0\to0$ being forbidden.
\end{enumerate}
The statistical weight of any level is $g=2J + 1$.\\

\n In addition to the forbidden lines, which break rule 4, there are the
intercombination, or \textbf{semi-forbidden} lines. These are transitions that
break any of the other rules. These transitions are about a million times lower
than the resonance transitions ($A\sim 10^2-10^3\ \mathrm{s^{-1}}$ compared to
$\sim10^8\ \mathrm{s^{-1}}$ for resonance lines). Forbidden lines are several
orders of magnitude less prevalent still, as they are magnetic dipole
transitions.\\

\n For forbidden or semi-forbidden transitions to be important, we must have
very low densities, since higher densities will make collisional processes
dominate over these weak transitions. As such, the emitted photons from
forbidden transitions are unlikely to be absorbed by another ion, so forbidden
line photons usually escape from a nebula and are thus very important coolants.
Examples of such transitions are [OIII] 4959, 5007 and [OII] 3726,29
\subsubsection{Hydrogen Line Series} % (fold)
\label{ssub:hydrogen_line_series}
The line transitions in a de-exciting electron (for instance, a recombination
electron that came in at high $n$, $\ell$ to form an excited hydrogen atom)
often produce a series of lines. Transitions that end at $n=1$ are called the
\textbf{Lyman Series}. Transitions from $n=2$ to $n=1$ are Lyman-$\alpha$
transitions, and those from $n=3$ to $n=1$ are Lyan-$\beta$ transitions, etc.
Similarly, transitions to the $n=2$ state form the Balmer series (denoted
H$\alpha$, H$\beta4$, etc.), and the transitions to $n=3$ form the Paschen
series.
% TODO transition from $n=2, \ell=0$ to $n=1, \ell=0$
% subsubsection hydrogen_line_series (end)
% subsection resonance_lines_and_selection_rules (end)
% section collisionally_excited_lines (end)
\section{Molecular Spectra} % (fold)
\label{sec:molecular_spectra}
\subsubsection*{Suggested Refeferences} % (fold)
\label{ssub:suggested_refeferences}
\begin{itemize}
  \item Draine Chapter 5
\end{itemize}
% subsubsection suggested_refeferences (end)
Molecules present additional complication since rotational and vibrational
energies become important. We'll first explore rotating molecules.
\subsection{Rotating Molecules} % (fold)
\label{sub:rotating_molecules}
Like nearly all forms of atomic energy, the energies of rotating molecules are
quantized to energy levels. These rotational energy levels are related tot he
moments of inertia of the molecules along the various axes of symmetry.

\paragraph{Eaample: Linear Diatomic Molecules} % (fold)
\label{par:eaample_linear_diatomic_molecules}
Suppose we have a linear diatomic molecule with characteristic lengthscale $r$
and reduced mass $\mu$. Then the moment of inertia of this molecule is
\begin{equation}
  \label{eq:ldmol:1} I = \mu r^2
\end{equation}
And the quantized rotational energy levels are
\begin{equation}
  \label{eq:ldmol:2} E = J(J+1)\frac{h^2}{8\pi^2 I}
\end{equation}
% TODO Finish this. This blackboard is in the worst position possible
% paragraph eaample_linear_diatomic_molecules (end)
% subsection rotating_molecules (end)
\subsection{Vibrating Molecules} % (fold)
\label{sub:vibrating_molecules}
% TODO Again, can't read for shit.
% subsection vibrating_molecules (end)
\subsection{Molecular Hydrogen} % (fold)
\label{sub:molecular_hydrogen}
H$_2$ does not radiate strongly because rotational transitions require a
heterogeneous linear molecule like CO. In the H$_2$ molecule, transitions occur
via electric quadrupole interaction. The least energetic transition is $J=0$ to
$J=2$. Thus, lifetiems of excited states are much, much longer than for the
ions (e.g.\ about 1000 years for the $J=2$ level). Hence, the rotational elvels
are populated by collisions.
% subsection molecular_hydrogen (end)
% section molecular_spectra (end)
\section{Transition Rates} % (fold)
\label{sec:transition_rates}
We return now to our two-level atom with two states, 1, and 2, separated by an
energy gap $E_{12}$. We can characterize the rate of decays from state 2
through some characteristic $e$-folding time via
\begin{equation}
  \label{eq:trans:1} N_2 = N_2(0)e^{-A_{21}t}
\end{equation}
Here, $A_{21}$ is proportional to the electric dipole matrix element, or
\begin{equation}
  \label{eq:trans:2} A_{21}\propto e^2\norm{\avg{\psi_1\vert\psi_2}}^2
\end{equation}
An atom with an electron in state 1 has a probabliity of absorbing a photon
given by
\begin{equation}
  \label{eq:trans:3} p_{12} = B_{12} u(\nu_{12})
\end{equation}
where $B_{12}$ is the Einstein absorption coefficient and $u(\nu)$ is the
energy density in radiation at a given frequency. We usually derive this energy
density from the \textbf{specific intensity}, $I_\nu$ ($I_\lambda$), which has
units of energy per area per time per sold angle per frequency (wavelength).
The specific intensity is a conserved quantity along a ray of light. To convert
this intensity to a flux through some surface, we must integrate this quantity
(really, the portion that is normal to the surface) over all incident angles, or
\begin{equation}
  \label{eq:trans:4} F_\nu = \int I_\nu\cos\theta\,d\Omega
\end{equation}
To get an energy density from the specific intensity (a process we'll discuss
later), we use
\begin{equation}
  \label{eq:trans:5} u_\nu = \frac{4\pi}{c}I_\nu
\end{equation}
Similarly, the probability for stimulated emission to occur is
\begin{equation}
  \label{eq:trans:6} p_{21} = B_{21}u(\nu_{12})
\end{equation}
where the two Einstein coefficients are related via
\begin{equation}
  \label{eq:trans:7} g_1 B_{12} = g_2B_{21}
\end{equation}
Using LTE, we can derive the spontaneous emission Einstein coefficient to be
% TODO get from Draine Ch. 6

% section transition_rates (end)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Wednesday, January 22, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Radiative Transfer} % (fold)
\label{sec:radiative_transfer}
\textit{Wednesday, January 22, 2014}
\subsection{Basic Definitions} % (fold)
\label{sub:basic_definitions}
\textbf{Energy flux} is the amount of radiative energy that passing through
some surface per unit area per unit time. This is in contrast to the
\textbf{specific intensity}, which is the amount of energy per unit time per
unit area per unit solid angle per unit frequency (or sometimes per unit
wavelength). Then the total energy through some bundle of rays is given by
\begin{equation}
  \label{eq:rt:1} dE = I_\nu\,dA\,dt\,d\Omega\,d\nu
\end{equation}
There are several quantities related to the specific intensity
\begin{enumerate}
  \item[(0)] \textbf{Mean Intensity}, defined by
  \begin{equation}
    \label{eq:rt:2} J_\nu \equiv \frac{1}{4\pi}\int I_\nu\,d\Omega
  \end{equation}
  which is just the specific unit averaged over all solid angles. Note that the
  differential of solid angle is
  \begin{equation}
    \label{eq:rt:3} d\Omega = \sin\theta\,d\theta\,d\phi
  \end{equation}
  This can be thought of as the zeroth moment of the specific intensity.
  \item[(1)] Correspondingly, the first moment of the specific intensity is the
  \textbf{specific flux}, given by
  \begin{equation}
    \label{eq:rt:4} F_\nu \equiv \int I_\nu \cos\theta\,d\Omega
  \end{equation}
  This is essentially the ``energy moment'' of the specific intensity.
  \item[(2)] While the first moment was an energy flux, the second moment is
  the momentum/pressure moment. Recalling that for a photon, $p = E/c$, we can
  get a momentum flux via
  \begin{equation}
    \label{eq:rt:5} P_\nu = \frac{1}{c}\int I_\nu \cos^2\theta\,d\Omega
  \end{equation}
  This is more familiarly known as the \textbf{radiation pressure}.
\end{enumerate}
As an example, consider a spaceship some distance $r$ from a (blackbody) star
of radius $R$ and effective temperature $T$. We'd like to know what flux the
spaceship observes from the star. Naively we may use the basic equation
\eqref{eq:rt:4} to start:
% TODO Add diagram
\begin{align}
  \label{eq:rt:6} F_\nu &= \int B_\nu(T)\cos\theta\sin\theta\,d\theta\,d\phi\\
  \label{eq:rt:7} &= B_\nu(T) \int_0^{2\pi} d\phi\int_0^{\theta_{\mathrm{max}}}
  \cos\theta\sin\theta\,d\theta\\
  \label{eq:rt:8} &= \pi B_\nu(T)\left(\frac{R}{r}\right)^2
\end{align}
where $\theta_{\mathrm{max}}$ is half the angular diamter of the star as seen
from the spaceship, $\theta_{\mathrm{max}} = \sin^{-1}(R/r)$.
% subsection basic_definitions (end)
\subsection{Equation of Radiative Transfer} % (fold)
\label{sub:equation_of_radiative_transfer}
% TODO Add diagram
Consider a source of radiation shining through some medium to an observer. The
change in the specific intensity, in its most general, is given by
\begin{equation}
  \label{eq:rt:9} dI_\nu = -\kappa_\nu I_\nu\,ds + j_\nu\,ds
\end{equation}
Where $j_\nu$ is the \textbf{emissivity}, which tells us the rate of emission
of radiation of that frequency per unit volume per unit solid angle, or
\begin{equation}
  \label{eq:rt:10} j_\nu \equiv \frac{1}{4\pi} n_u A_{u\ell}h\nu
\end{equation}
In addition to the emissivity, we must also consider the \textbf{attenuation},
$\kappa_\nu$, which measures the rate at which radiation is absorbed, which is
obviously proportional to the incident radiation, $I_\nu$. The attenuation has
an absorption term and an emission term:
\begin{align}
  \label{eq:rt:11} \kappa_\nu &= n_\ell\sigma_{\ell u}(\nu) -
  n_u\sigma_{u\ell}(\nu)\\
  \label{eq:rt:12} &= n_\ell \sigma_{\ell u}(\nu)\left[1 - \frac{n_u \sigma_{u
  \ell}}{n_\ell \sigma_{\ell u}}\right]\\
  \label{eq:rt:13} &= n_\ell \sigma_{\ell u}(\nu)\left[1 -
  \frac{n_u/n_\ell}{g_u/g_{\ell}}\right]
\end{align}
where in the last step we've used detailed balance to eliminate cross sections
in favor of statistical weights. Now in LTE, the number densities of states are
given by a Boltzmann distribution:
\begin{equation}
  \label{eq:rt:14} \frac{n_u}{n_\ell} = \frac{g_u}{g_\ell} e^{-h\nu/kT}
\end{equation}
So for sufficiently high frequencies ($h\nu \gg kT$), the second term in
\eqref{eq:rt:13} can be neglected, giving us simply
\begin{equation}
  \label{eq:rt:15} \kappa_\nu \approx n_\ell \sigma_{\ell u}
\end{equation}
This approximation is excellent for the vast majority of astrophysical
applications. A notable exception is masers, where stimulated emission is
important (note that these are at low wavelengths; microwaves in fact).
\subsubsection{Example: Lyman Alpha Forest from AGNs} % (fold)
\label{ssub:example_lyman_alpha_forest_from_agns}
% TODO add diagram
Quasars at $z\approx 3$ emit strong lymnan alpha radiation that travels through
a vast expanse of IGM. As it goes through, lyman alpha opacity robs the
incident spectrum of lyman alpha radiation. The transition back to the ground
state is a permitted transition, so it happens fast and emits photons
isotropically. Thus, very few of the absorbed photons are re-emiited in the
original direction. Assuming the emissivity vainishes along this path (no
spontaneous production of Lyman alpha photons), we get
\begin{equation}
  \label{eq:rt:16} \frac{dI_\nu}{I_\nu} = -\kappa_\nu\,ds \qquad \Rightarrow
  \qquad \left. \ln I_\nu\right\vert_s^{\mathrm{obs}} =
  -\int_s^{\mathrm{obs}}\kappa_\nu\,ds
\end{equation}
which inspires us to define the \textbf{optical depth}, a dimensionless
parameter characterizing the relative opaqueness of a path, via
\begin{equation}
  \label{eq:rt:17} \tau_\nu = \int ds\, \kappa_\nu
\end{equation}
this gives us another way to express \eqref{eq:rt:16}:
\begin{equation}
  \label{eq:rt:18} I_\nu = I_\nu(s) e^{-\tau_\nu(\mathrm{obs})}
\end{equation}
% subsubsection example_lyman_alpha_forest_from_agns (end)
\subsubsection{Optical Depth} % (fold)
\label{ssub:optical_depth}
Thus there are two limiting cases, the optically thin case, where $\tau \ll 1$
and the optically thick $\tau \gg 1$ case. A source viewed through an optically
thin medium will suffer very little attenuation, whereas a source viewed
through an optically thick medium will be almost totally attenuated.\\

\n With the optical depth defined, we can recast the equation of radiative
transfer neatly in terms of it:
\begin{equation}
  \label{eq:rt:19} dI_\nu + I_\nu d\tau_\nu = \frac{j_\nu}{\kappa_\nu}d\tau_\nu
\end{equation}
to make things work out nicely, we multiply both sides of \eqref{eq:rt:19} by
the integrating factor
$\exp\left(\int_{\mathrm{obs}}^\ell\kappa_\nu\,ds\right)$ to get
\begin{equation}
  \label{eq:rt:20} \int_s^{\mathrm{obs}}d\left[e^{\int_{\mathrm{obs}}^\ell
  \kappa_\nu ds}I_\nu\right] = \int_0^{\tau_\nu} \frac{j_\nu}{\kappa_\nu}
  e^{\int_{\mathrm{obs}}^\ell \kappa_\nu ds}d\tau
\end{equation}
With some more jiggery-pokery, we get
\begin{equation}
  \label{eq:rt:21} \boxed{I_\nu(\mathrm{obs}) = I_\nu(s)e^{-\tau_\nu} +
  \int_0^{\tau_\nu} \frac{j_\nu}{\kappa_\nu} e^{-(\tau_\nu - \tau_\nu')}\
  d\tau_\nu'} \end{equation}
The first term just give the attenuated original specific intensity, and the
second term adds in the sources along the path, each attenuated from its source
position to the observer.\\

\n Now if the cloud in the medium is in LTE, where the photon emission rate is
equal to the photon absorption rate, we have the emissivity of
\begin{equation}
  \label{eq:rt:22} j_\nu \equiv \kappa_\nu B_\nu
\end{equation}
If we are at a uniform tepmerature, $T$, the equation of radiative transfer
becomes
\begin{equation}
  \label{eq:rt:23} I_\nu(\mathrm{obs}) = I_\nu(s)e^{-\tau_\nu} +
  B_\nu(T)\int_0^{\tau_\nu}e^{-(\tau_\nu - \tau_\nu')}d\tau'
\end{equation}
Now this is an integral we can do, so this simplifies to
\begin{equation}
  \label{eq:rt:24} \boxed{I_\nu(\mathrm{obs}) = I_\nu(s) e^{-\tau_\nu} +
  B_\nu(T) \left(1- e^{-\tau_\nu}\right)}
\end{equation}
If the very diffuse limit, $\tau_\nu \ll 1$, we may take first term expansions
of the exponentials to get simply
\begin{equation}
  \label{eq:rt:25} I_\nu(\mathrm{obs}) = I_\nu(s)(1-\tau_\nu)
\end{equation}
And for the very thick cloud, where $\tau_\nu \gg 1$, the exponentials vanish,
giving simply
\begin{equation}
  \label{eq:rt:26} I_\nu(\mathrm{obs}) = B_\nu(T)
\end{equation}
% subsubsection optical_depth (end)
\subsubsection{Radio Astronomy} % (fold)
\label{ssub:radio_astronomy}
For radio astronomy, we are typically in the regime where $h\nu\ll kT$, so
stimulated emission is important. Radio astronomers often speak of a
\textbf{brightness temperature}, which is the temperature a source would need
to have the observed brightness at a particular frequency (this temperature
need not be the same for each frequency). Inverting the Planck function, we can
get an analytic expression for the brightness temperature:
\begin{equation}
  \label{eq: rt:26} T_B(\nu) = \frac{h\nu}{k} \ln\left[\frac{2h\nu^3}{c^2
  I_\nu} + 1\right]{-1}
\end{equation}
Radio astronomers will also use an antennae temperature via
% TODO Explain this better
\begin{equation}
  \label{eq:rt:27} \boxed{T_A \equiv \frac{c^2}{2k\nu^2}I_\nu}
\end{equation}
The transfer equation becomes
\begin{equation}
  \label{eq:rt:28} \frac{c^2}{2\nu^2k}I_\nu = \frac{c^2}{2\nu^2 k}I_\nu(s) +
  \frac{B_\nu c^2}{2\nu^2k} \left(1-e^{-\tau_\nu}\right)
\end{equation}
or simply
\begin{equation}
  \label{eq:rt:29} T_A(\mathrm{obs}) = T_A(s) + T_c\left(1-e^{-\tau_\nu}\right)
\end{equation}
In the limit where the antennae temperature of the source is much greater than
the cloud temperature and the cloud is optically thin, \eqref{eq:rt:29}
simplifies to
\begin{equation}
  \label{eq:rt:30} T_A = T_{A,s}(1-\tau_\nu) + T_c\left(1-(1-\tau_\nu)\right) =
  T_{A,s} - (T_{A,s} - T_c)\tau_\nu
\end{equation}
In the optically thick limit, we just get
\begin{equation}
  \label{eq:rt:31} T_A(\mathrm{obs}) = T_c
\end{equation}
% subsubsection radio_astronomy (end)
% subsection equation_of_radiative_transfer (end)
\subsection{Resonance Line Transfer} % (fold)
\label{sub:resonance_line_transfer}
\textbf{MISSING: MONDAY, JANUARY 27, 2014 (EQUIVALENT WIDTH, CURVE OF GROWTH,
ETC.)}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WEDNESDAY, JANUARY 29, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\n \textit{Wednesday, January 29, 2014}

\subsubsection{The Voigt Profile} % (fold)
\label{ssub:the_voigt_profile}
The equivalent width, in wavelength units (though not per unit wavelength), is
given by
\begin{equation}
  \label{eq:ew:1} W_\lambda = \frac{\lambda^2}{c} \int (1-e^{-\tau_\nu}d\nu)
\end{equation}
where the optical depth is
\begin{equation}
  \label{eq:ew:2} \tau_nu = s\,N\,\phi(\nu)
\end{equation}
and $\phi(\nu)$ is the line profile.\\

\n From quantum mechanics, we remember that the width of a line is inversely
proportional to the lifetime of the state via the uncertainty principle,
$\Delta E \Delta t \sim h$. This natural shape is the Lorentzian, which is like
a sharper Gaussian with stronger wings. In functional form, the natural line
profile is then
\begin{equation}
  \label{eq:ew:3} \phi_\nu = \frac{4\gamma_{21}}{16\pi^2(\nu - \nu_{21})^2 +
  \gamma_{21}^2}
\end{equation}
For a line generated by a transition from state 2 to state 1. To get the width
of this line, let's compute the half-width at half maximum (HWHM), or the
frequency $\nu_{\mathrm{HM}}$ where $\phi(\nu_{\mathrm{HM}}) = 2/\gamma_{21}$:
\begin{equation}
  \label{eq:ew:4} \frac{4\gamma_{21}}{16\pi^2\Delta\nu^2+\gamma-{21}^2} =
  \frac{2}{\gamma_{21}} \qquad \Rightarrow \qquad \Delta \nu =
  \frac{\gamma_{21}}{4\pi}
\end{equation}
So the full width at half maximum (FWHM) is
\begin{equation}
  \label{eq:ew:5} \boxed{2\Delta \nu = \frac{\gamma_{21}}{2\pi}}
\end{equation}
For example, for the Hydrogen $2\to 1$ transition, we can look up $\gamma_{21}
= 6.265\times 10^8\ \mathrm{s^{-1}}$, which corresponds to a line width or
velocity of
\begin{equation}
  \label{eq:ew:6} \Delta v = \frac{\Delta\nu}{\nu_{21}}c = 0.0121\ \mathrm{km\
  s^{-1}} \left[\frac{\lambda_{21}\gamma_{21}}{7618\ \mathrm{cm\ s^{-1}}}\right]
\end{equation}
So this is a \emph{very} thin width, especially considering that velocities due
to thermal motions are typically much higher than this velocity dispersion. The
true line profile then is given by a convolution of the particle velocity
distribution and the natural Lorentizan distribution. In general, the velocity
distribution of the particles is a Gaussian, and the convolution of a Gaussian
and a Loretnzian distribution is called a \textbf{Voigt Profile}. Analytically,
it looks like
\begin{equation}
  \label{eq:ew:7} \boxed{\phi_\nu^{\mathrm{Voigt}} = \frac{1}{\sqrt{2\pi}}\int
  \frac{dv}{\sigma_{v}}e^{-v^2/(2\sigma^2)}\frac{4\gamma_{21}}{16\pi^2\left[\nu-
  (1-v/c)\nu_{21}\right]^2 + \gamma_{21}^2}}
\end{equation}
Unfortunately, there is no analytic solution for this integral. Instead, we'll
look at two regimes within the Voigt profile.
\paragraph{Near Line Center (Gaussian)} % (fold)
\label{par:near_line_center_gaussian_}
Near the line center, where the fraction in \eqref{eq:ew:7} is nearly constant,
the profile appears to be a gaussian:
\begin{equation}
  \label{eq:ew:8} s = \sqrt{\pi} \frac{e^2}{m_e c} \frac{f_{12}\lambda_{21}}{b}
  e^{-v^2/b^2}
\end{equation}
where we've used the \textbf{Doppler Parameter} $b=\sqrt{2}\sigma$. 
% paragraph near_line_center_gaussian_ (end)

\paragraph{Wings} % (fold)
\label{par:wings}At large $\Delta \nu$, we get
\begin{equation}
  \label{eq:ew:9} s = \sqrt{\pi} \frac{e^2}{m_ec}\frac{f_{12}\lambda_{12}}{b}
  \left[\frac{1}{4\pi^{3/2}}\frac{\gamma_{21}\lambda_{21}}{b}\frac{b^2}{v^2}
  \right]
\end{equation}
% paragraph wings (end)

\paragraph{Transition} % (fold)
\label{par:transition}
The transition between these two regimes occurs around 
\begin{equation}
  \label{eq:ew:10} e^{z^2}=r\pi^{3/2} \frac{bz^2}{\gamma_{21}\lambda_{21}}
\end{equation}
where $z\equiv v/b$. For Ly\,$\alpha$, this corresponds to $z = 3.2$, or
$\Delta v = 32\ \mathrm{km\ s^{-1}}(b/10\ \mathrm{km\ s^{-1}})$. Remember that
for a Gaussian, the FWHM is $2\sqrt{2\ln\sigma}$\\
% paragraph transition (end)

\n Now if the absorbing medium is optically thick at a certain frequency (i.e.,
$\tau_nu > 1$, then define $x\equiv v/b$ and again $v = (\nu -
\nu_{21})/\nu_{21}$). Define $x_1$ such that the optical depth is unity at that
frequency. The optical depth dies off exponentially as we move away from the
resonance frequency,
\begin{equation}
  \label{eq:ew:11}\tau_\nu = \tau_0 e^{-x^2}
\end{equation}
Solving for $x_1$ then gives us
\begin{equation}
  \label{eq:ew:12} x_1 = \sqrt{\ln \tau_0}
\end{equation}
Taking a Taylor expansion of the attenuation, we find
\begin{equation}
  \label{eq:ew:13} 1-e^{-\tau_\nu} \approx \begin{cases}
    1, &\text{ for }\abs{x} < x_1\\
    \tau_\nu, &\text{ for }\abs{x} \geq x_1
  \end{cases}
\end{equation}
Then the equivalent width in units of wavelength is
\begin{align}
  \label{eq:ew:14} \frac{W(\lambda)}{\lambda_{12}} &= 2\times
  \frac{\lambda_{12}}{c} \frac{\nu_{12}b}{c} \int_0^\infty
  (1-e^{-\tau_\nu})\,d\nu\\
  \label{eq:ew:15} &\approx \frac{2b}{c}\left[ \int_0^{x_1}dx + \int_0^\infty
  \tau_0e^{-x^2}dx - \int_0^{x_1}\tau_0e^{-x^2}dx\right]\\
  \label{eq:ew:16} &\approx \frac{2b}{c} \left[\sqrt{\ln\tau_0} +
  \tau_0\frac{\sqrt{\pi}}{2}\left\{1 -
  \mathrm{erf}\left(\sqrt{\ln\tau_0}\right)\right\}\right]
\end{align}
In thte optically thick limit, where $\tau_0\to\infty$, we may ignore the
bracketed term, giving us simply
\begin{equation}
  \label{eq:ew:17} \boxed{\frac{W_\lambda}{\lambda} =
  \frac{2b}{c}\sqrt{\ln\tau_0}}
\end{equation}
where
\begin{equation}
  \label{eq:ew:18} \tau_0 = \frac{s N_1 \lambda_{12}}{\sqrt{\pi} b}
\end{equation}
Scaling to the Ly $\alpha$ line, 
\begin{equation}
  \label{eq:ew:19} \tau_\nu = 1 \qquad \Rightarrow \qquad \boxed{N = 8\times
  10^{13}\ \mathrm{cm^{-2}}\frac{b}{10\ \mathrm{km\ s^{-1}}}\frac{1}{f}
  \frac{1216\,\mathrm{\AA}}{\lambda_{12}}}
\end{equation}
So the ``Lyman Alpha Forest'' clouds are all optically thin with column
densities $N(\mathrm{H}\,I) < 10^{14}\ \mathrm{cm^{-2}}$. Then there are Lyman
limit systems, which are optically thick. Then finally, there are the Damped Ly
$\alpha$ Absorbers, or DLAs, which we explore now.\\

\n For \emph{very} optically thick clouds, where $\tau_0 > 10^4$, the
attenuation goes as
\begin{equation}
  \label{eq:ew:20} 1-e^{-\tau} \approx \begin{cases}
    1, &\text{ if }\abs{x}\leq x_1\\
    \tau_0\frac{a}{\sqrt{\pi}}x^{-2}, &\text{ if }\abs{x}> x_1
  \end{cases}
\end{equation}
where
\begin{equation}
  \label{eq:ew:21} a = \frac{\gamma_{12}c}{b \nu_{12}}
\end{equation}
Then the equivalent width is
\begin{equation}
  \label{eq:ew:22} \boxed{\frac{W_\lambda}{\lambda} = \frac{2}{c}
  \sqrt{\gamma_{12}}{\lambda_{12}^2sN_1}}
\end{equation}
Again scaling to Ly $\alpha$, we get
\begin{equation}
  \label{eq:ew:23} W_\lambda = 7.3\ \mathrm{\AA}\
  \sqrt{\frac{N(\mathrm{H\,I})}{10^{20}\ \mathrm{cm^{-2}}}}
\end{equation}
And now we've identified the three main modes in the curve of growth at small,
intermediate, and high column depths.
% subsubsection the_voigt_profile (end)
\subsubsection{21 cm Line} % (fold)
\label{ssub:21_cm_line}
In the hydrogen atom, the electron-proton spin coupling causes a hyperfine
splitting between the two spin states of the electron in neutral hydrogen. The
photon emitted through a spin-flip transition has a wavelength of 21 cm. In
temperature units, this corresponds to a 0.06816 K. Recalling that the CMB
temperature is $T_{\mathrm{CMB}} = 2.7\ \mathrm{K}$, the Boltzmann distribution
of the two spin states,
\begin{equation}
  \label{eq:21cm:1} \frac{n_2}{n_1} = \frac{g_2}{g_1} e^{-h\nu_{12}/kT}
\end{equation}
means that the distribution is largely determined by the statistical weights.
Thus, $n_2/n_1 = 3$ pretty much always. Thus
\begin{equation}
  \label{eq:21cm:2} \frac{n_2}{n(\mathrm{H\,I})} = \frac{3}{4}; \qquad
  \frac{n_1}{n(\mathrm{H\,I})} = \frac{1}{4}
\end{equation}
The emissivity in the 21-cm line is
\begin{equation}
  \label{eq:21cm:3} \gamma_{21} = \frac{n_2 A_{21} h\nu_{21}}{4\pi} \phi(\nu) =
  \frac{3}{16\pi} n(\mathrm{H\,I})A_{21}h\nu_{21}\phi(\nu)
\end{equation}
And the absorption is
\begin{align}
  \label{eq:21cm:4} \kappa_\nu =& n_1\sigma_{12} - n_2\sigma_{21}\\
  \label{eq:21cm:5} &= n_1 \frac{g_2}{g_1} \frac{A_{21}}{8\pi} \lambda_{21}^2
  \left[1 - e^{-h\nu_{21}/(kT)}\right]\phi(\nu)\\
  \label{eq:21cm:6} &\approx \frac{3}{32\pi} A_{21}\frac{hc \lambda_{12}}{kT}
  n(\mathrm{H\,I})\phi(\nu)
\end{align}
And the optical depth is
\begin{equation}
  \label{eq:21cm:7} \tau_\nu = \int_s^o\kappa_\nu\,ds
\end{equation}
For a Gaussian $\phi(\nu)$, we get
\begin{equation}
  \label{eq:21cm:8} \tau_\nu = 2.190 \frac{N(\mathrm{H\,I})}{10^{21}\
  \mathrm{cm^{-2}}}\left[\frac{100\
  \mathrm{K}}{T}\right]\left[\frac{\mathrm{km\ s^{-1}}}{b/\sqrt{2}}\right]e^{-v^
  2/b^2}
\end{equation}
% subsubsection 21_cm_line (end)
% subsection resonance_line_transfer (end)
% section radiative_transfer (end)
\section{Additional Microphysics} % (fold)
\label{sec:additional_microphysics}
\subsection{Emission and Absorption by a Thermal Plasma} % (fold)
\label{sub:emission_and_absorption_by_a_thermal_plasma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MONDAY, FEBRUARY 3, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textit{Monday, February 3, 2014}\\

\n Suppose we have a thermal (i.e. Maxwellian) plasma at temperature $T\sim
8000\ \mathrm{K}$. The continuum spectrum of such a plasma is \emph{not} simply
a blackbody since such a plasma is not necessarily optically thick. Rather, a
collection of sources of continuum sources make a rich spectrum. Such sources
include recombination (free electrons recombine with ions, giving off an
ionization potential plus their kinetic energy into a photon), the two-photon
continuum, where an electron undergoes a forbidden transition by giving off two
photons that preserve spin properly but whose energies add up to the difference
in potential. We'll discuss these processes in more detail now.\\

\n \textbf{Bound-bound} transitions are those where an electron starts being
bound and ends being bound. One such example is cooling via [C\,II] 158 $\mu$m,
which is cooling via collisional excitation. The actual volumetric cooling rate
is simply given by
\begin{equation}
  \label{eq:bbound:1} \frac{\mathrm{energy\
  radiated}}{(\mathrm{time})(\mathrm{volume})} = n_{\mathrm{C^+}}n_{\mathrm{H}}
  h\nu \alpha_{12} = n_{\mathrm{C^+}}n_{\mathrm{H}}(8.1\times
  10^{-24}\,\mathrm{erg\ s^{-1}\ cm^{-3}})
\end{equation}
where we can look up the recombination rate,
\begin{equation}
  \label{eq:bbound:2} \alpha_{21}(\mathrm{H-C^+})=8\times 10^{-10}\
  \mathrm{cm^3\ s^{-1}}
\end{equation}
and by detailed balance,
\begin{equation}
  \label{eq:bbound:3} \alpha_{12} = \alpha_{21} \frac{g_2}{g_1}e^{-92\
  \mathrm{K}/T}
\end{equation}
If we assume that hydrogen is mostly neutral and carbon is mostly singly
ionized and use typical abundances, we can scale this cooling rate to just the
hydrogen number density:
\begin{equation}
  \label{eq:bbound:3} \cdots = n_{\mathrm{H}}^2(3.24\times 10^{-27}\
  \mathrm{erg\ s^{-1}\ cm^{-3}})
\end{equation}
In \HI regions, this is a dominant source of cooling.\\
% TODO add other types of cooling

\n \textbf{Free-Free Emission and Absorption}
For free-free radition, when two charged particles scatter and give off a
photon, the power radiated per unit volume scales as
\begin{equation}
  \label{eq:ffree:1} P\propto T^{1/2} Z_i^2 n_i n_e
\end{equation}
most of this power is near $h\nu\sim kT$. The radio and microwave free-free
spectrum is almost flat, declining with increasing frequency as $\nu^{-0.121}$.
On the absorption side, the attenuation coefficient follows from Kirchhoff's
law: $\kappa_\nu = j_\nu/B_\nu(T)$. This attenuation is very strong at low
frequencies, but negligible in the ISM for $\nu > 10$ GHz. \HII regions can be
optically thick at $\nu < 1$ GHz.\\

\n \textbf{Bound-free emission}, or radiative recombination, is the process
where an ion recaptures an electron, liberating a photon with energy equal to
the ionization potential plus the kinetic energy of the electron. The
lowest-energy electrons are more likely to be captured by this process, and the
rate is thus dependent on the temperature of the gas. In this process, the ion
captures the electron, but the new less-ionized ion is probably still in an
excited state. This state will eventually decay to the ground state, emitting
one or more photons on the way down. The first photon contributes to the
recombination continuum, and the subsequent decays power the recombination
lines. Thus, bound-free processes ultimately give rise to strong continuum
\emph{and} line emission that depends on the atomic physics of the ion in
question.\\

\n \textbf{Photoionization} is the inverse process of radiative recombination,
where a bound electron absorbs a photon and escapes its potential well. For
photoionizing an electron, there is a well-known functional form of the
photoionization cross section:
\begin{equation}
  \label{eq:photoion:1} \sigma(\nu) \approx \sigma_0(Z)\nu^{-3};\qquad
  \sigma_0(Z) = 6.304\times 10^{-18}Z^{-2}\ \mathrm{cm^{-2}}
\end{equation}
This is only valid for hydrogenic (single electron) ions. It gets more
complicated with 3 or more electrons because the resulting ion can be left in a
variety of states. The absorption edge is at the minimum photon energy for
photon energy for photoionization from a shell. That is, there is a strong edge
for hydrogen at $h\nu = 13.6\ \mathrm{eV}$ below which the cross section
vanishes (these photons cannot photoionize the electron). Despite the low
abundances present, this can dominate at high energies.\\

\n We can relate the cross section for photoionization from a given level to
the cross section for recombination to that same level via the Milne Relation.
For Hydrogen, we have
\begin{align}
  \label{eq:photoion:2} \sigma &= 6.3042\times 10^{-18}\ \mathrm{cm^2}\
  \left(\frac{\nu}{\nu_0}\right)^{-3.5}\\
  \label{eq:photoion:3} \alpha &= 4.18\times 10^{-13}\left(\frac{T}{10^4\
  \mathrm{K}}\right)^{-0.72}\ \mathrm{cm^3\ s^{-1}}
\end{align}
  \textbf{X-Ray Fluorescence} is when one electron drops down (from the L
  shell) to fill the vacancey (K shell), so the atom emits a photon (Ka).
  Similarly, Auger ionization is when one electron drops down to fill the hole
  left by X-Ray ionization from the K shell. A second electron is promoted to
  an excited (and unbound) level. This looks like
  \begin{equation}
    \label{eq:x-ray} X + h\nu \to (X^+)^* + e^- \to X^{+n} + ne^- \qquad (n
    \geq 2)
  \end{equation}
  Thus, two higher-energy electrons are produced. The inverse of this process
  is called \textbf{di-electronic recombination}. Di-electronic recombination
  is important in very high-temperature plasmas. When it is important, it often
  dominates over regular radiative recombination. The electron rarely has
  enough energy at low temperature to reproduce a doubly excited state. However
  Mg\,II and C\,III are exceptions in which the dielectronic recombination rate
  is significant at $10^4\ \mathrm{K}$. Additiaonlly, this can be important at
  low temperatures if one of the excited states is a fine-structure state, like
  C\,II $\to$ C\,I and others\\
  % TODO missed a bullet point here

  \n \textbf{Collisional ionization} is very much like the collisional
  excitation we've talked about before, but enough energy is transferred to
  fully ionize an electron into the free continuum. There are other processes
  we haven't covered like three-body processes, charge-exchange, cosmic ray
  ionization, dissociateive recombination, and neutralization by grains.

% subsection emission_and_absorption_by_a_thermal_plasma (end)
\subsection{Collisional Ionization Equilibrium} % (fold)
\label{sub:collisional_ionization_equilibrium}
  In a state of collisional ionization equilibrium, the stages of ionization
  are coupled by the rates for collisional ionization, recombination, and\ldots
  % TODO missed a ton here
  
  \n All this information can be combined to get a colling function, $Q(n_e, T,
  Z)=\Lambda n_e n$, which describes the total energy lost by a plasma per unit
  volume per unit time.
% subsection collisional_ionization_equilibrium (end)
% section additional_microphysics (end)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Wednesday, February 5, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{H II Regions} % (fold)
\label{sec:h,ii_regions}
\textit{Wednesday, February 5, 2014}
\subsection{Thermal Equilibrium} % (fold)
\label{sub:thermal_equilibrium}
In a region of gas around a bright star, a pocket of ionized hydrgen can form,
driven by the photoionizing radiation from the star. This region is called an
\textbf{H\,II Region}. These are usually in thermal equilibrium, where the
heating rate is equal to the cooling rate. Heating sources include
photoionization from the star, cosmic rays, and collisional de-excitation. The
important cooling sources are recombination, free-free processes, and
collisional excitation. If the gas is in \textbf{ionization equilibrium}, the
photoionization and recombination rates are equal as well. That is, the
ionization fraction of a given species is constant in time. This means that
these two heating/cooling rates are also equal.\\

\n The mean energy of a photo-electron is simply
\begin{equation}
  \label{eq:h2:1} \avg{E_{\mathrm{pe}}} = \frac{1}{2}m_e\avg{v_{\mathrm{pe}}^2}
\end{equation}
However the mean energy of recombining electrons,
\begin{equation}
  \label{eq:h2:2} \avg{E_{\mathrm{recomb}}} =
  \frac{1}{2}m_e\avg{v_{\mathrm{recomb}}^2}
\end{equation}
is always lower. Thus, these rates being equal actually introduces a net
heating process:
\begin{equation}
  \label{eq:h2:3} \Delta E = \frac{1}{2}m_e\left(\avg{v_{\mathrm{pe}}^2} -
  \avg{v_{\mathrm{recomb}}^2}\right)
\end{equation}
This heat source must be balanced by some combination of free-free emission and
collisional excitation.
\subsubsection{Photoionization Heating} % (fold)
\label{sub:photoionization_heating}
The energy input by photoionization is determined from the stellar spectrum,
$J_\nu$:
\begin{equation}
  \label{eq:h2:4}\textrm{Heating Rate} = \int_{\nu_0}^\infty \frac{4\pi
  J_\nu}{h\nu}\sigma_{\mathrm{ph}}(\nu)n_{\mathrm{H}}h(\nu - \nu_0)\,d\nu
\end{equation}
where $\nu_0$ is the ionization potential of Hydrogen (13.6 eV/$h$). The
ionization rate is just this without the excess energy term:
\begin{equation}
  \label{eq:h2:5}\textrm{Ionization Rate} = \int_{\nu_0}^\infty \frac{4\pi
  J_\nu}{h\nu}\sigma_{\mathrm{ph}}(\nu)n_{\mathrm{H}}\,d\nu
\end{equation}
The average photoelectron energy is just the ratio of these two quantities:
\begin{equation}
  \label{eq:h2:6}\textrm{Mean Photoelectron Energy} = \frac{\textrm{Heating
  Rate}}{\textrm{Ionization Rate}}
\end{equation}
If we assume a blackbody spectrum,
\begin{equation}
  \label{eq:h2:7} 4\pi J_\nu = \pi B_\nu\propto \frac{\nu^3}{e^{h\nu/kT} - 1}
  \propto \nu^3 e^{-h\nu/kT}
\end{equation}
where we are assuming that the ionization frequencies are on the Rayleigh Jeans
tail. The ``energy'' of the typical photosphere photon is
\begin{equation}
  \label{eq:h2:8} kT_{\mathrm{eff}} = 2.6\ \mathrm{eV}\
  \frac{T_{\mathrm{eff}}}{30\,000\ \mathrm{K}}
\end{equation}
Carrying out all the math (not shown here), we find an approximate form for the
average photoelectron energy to be
\begin{equation}
  \label{eq:h2:9} E_{\mathrm{pe}}=\avg{\frac{3}{2}kT_i}\approx
  kT_{\mathrm{eff}}\left[1-\frac{kT_{\mathrm{eff}}}{h\nu_0}\right]
\end{equation}
This is roughly linear at low temperatures since the second term is negligible.
However, when $kT_{\mathrm{eff}}$ becomes comparable to $h\nu_0$, at around 30\
000 K, the slope begins to taper off. That is, there is a decreasing return on
simply cranking up the temeprature of the star.\\

\n For an optically thick (at line edge) region, the low energy region of the
stellar spectrum (near $h\nu_0$) is effectively removed by the inner regions of
the cloud. Thus, at larger distances, there are fewer ionizations (both due to
the exponential die-off of higher energy photons in the spectrum and due to the
$\nu^{-3}$ dependence of the ionization cross section). However, those
ionizations that do happen are at higher energy, so the man photo-electron
energy is actually higher at larger distances from the source.
% subsection photoionization_heating (end)
\subsubsection{Cooling By Recombination} % (fold)
\label{sub:cooling_by_recombination}
The cooling by recombination is given roughly by
\begin{equation}
  \label{eq:h2:10} L_{\mathrm{R}} \propto \frac{1}{2}\avg{v_{\mathrm{R}}^2}
  n_pn_e \alpha
\end{equation}
where, remember, $\alpha$ has dimensions of $(\mathrm{volume})
(\mathrm{time})^{-1}$. $\alpha$ is actually a more complicated term than we
might expect:
\begin{equation}
  \label{eq:h2:11} \alpha = \sum_n\sum_L\alpha_{\mathrm{n,L}}
\end{equation}
Where the sums are over the transitions from the various energy levels $n$ and
angular momentum states $L$. In general, $\alpha$ scales with the inverse
square root of temperature,
\begin{equation}
  \label{eq:h2:12} \alpha\propto T_e^{-1/2}
\end{equation}
and so (due to the thermal energy term in the cooling rate), the cooling rate
is proportional to the square root of temperature:
\begin{equation}
  \label{eq:hw:13} L_{\mathrm{R}} \propto \sqrt{T_e}
\end{equation}
Thus hotter gasses cool faster via recombination. A more complete, albeit
unclear statement of the cooling rate due to recombination is
\begin{equation}
  \label{eq:h2:14} L_{\mathrm{R}}(\mathrm{H}) = n_e n_p kT_e
  \beta_{B}(\mathrm{H}^0, T)
\end{equation}
where $\beta$ is now a recombination coefficient that we can look up as a
function of the ion in question and the electron temperature. The subscript $B$
refers to \textbf{Case B Recombination}. In fact, the coefficent $\beta_B$ is
more precisely given by
\begin{equation}
  \label{eq:h2:15} \beta_B =
  \sum_{n=2}^{\infty}\sum_{L=0}^{n-1}\beta_{nL(\mathrm{H}^0, T)}
\end{equation}
This means we can ignore all recombinatiosn directly to the ground state (the
first sum excludes $n=1$). The reason we can make this assumption is that in an
optically thick cloud, the resulting Lyman continuum photon from a direct
recombination will be absorbed so fast so as to effectively cancel out the
recombination/ionization event (the mean free path in the standard ISM is less
than a parsec). This approximation is known as the \textbf{On The Spot}
approximation (OTS). If the gas is \emph{not} optically thick, one must use
case A recombination approximations.
% subsection cooling_by_recombination (end)
\subsubsection{Collisional Cooling} % (fold)
\label{ssub:collisional_cooling}
Thus far we have only considered Hydrogen's contriubtion. However, heavier
elements like C, N, O, etc.\ are important coolants despite their relatively
low abundances. Recall that the collisional cross section as a function of
relative velocity for a given transition is
\begin{equation}
  \label{eq:h2:16} \sigma_{12}(v) = \frac{\pi
  \hbar^2}{m^2v^2}\frac{\Omega(1,2)}{\omega_1}
\end{equation}
And the cooling rate due to collisional excitation is then (in the low density
limit)
\begin{equation}
  \label{eq:h2:17}L_c= n_e n_1 h\nu_{12}q_{12}
\end{equation}
where $q_{12}$ is ascertained from $q_{21}$ and detailed balance:
\begin{equation}
  \label{eq:h2:18} q_{21} = \int_0^\infty \sigma_{21}(v)f(v)v\,dv =
  \frac{8.6\times 10^{-6}}{\sqrt{T}}\frac{\Omega(1,2)}{\omega_{2}}
\end{equation}
And so
\begin{equation}
  \label{eq:h2:19} q_{12} = q_{21} \frac{\omega_2}{\omega_1} e^{-\Delta E / kT}
\end{equation}
In the high density limit, 
\begin{equation}
  \label{eq:h2:20} q_{12}n_1n_e = n_2 A_{21} + n_2 n_e q_{21} \approx
\end{equation}
And so the cooling rate here is given by
\begin{equation}
  \label{eq:h2:21} L_c=n_2 h\nu_{21}A_{21} = n_1 n_e q_{12} h\nu_{12} \left[
  \frac{1}{1 + \frac{q_{21}n_e}{A_{21}}}\right]
\end{equation}
From the recombination and collisional cooling rates, we can get a cooling
rate. Then by matching the summed cooling rate to our heating rate due to
photoionization (and possibly collisional de-excitation), we can identify an
equilibrium temperature for the region.
% subsubsection collisional_cooling (end)
% subsection thermal_equilibrium (end)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MONDAY, FEBRUARY 10, 2014 %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Str\"omgren Spheres} % (fold)
\label{sub:stromgren_spheres}
\subsubsection{Ionization Equilibrium of a Pure H Nebula} % (fold)
\label{ssub:ionization_equilibrium}
If a pure hydrogen nebula is in ionization equilibrium, the rate of ionizations
equals the rate of recombinations:
\begin{equation}
  \label{eq:strom:1} \int_{\nu_0}^\infty \frac{4\pi J_\nu}{h\nu}
  \sigma(\nu)n_{\mathrm{H}}\,d\nu = \alpha(\mathrm{H}, T)n_e n_{\mathrm{H^+}}
\end{equation}
For now we assume that the nebula is large enough that all of the ionizing
photons are used in keeping the \HII region ionized. What is unclear is whether
the edge at which the neutral fraction goes up is sharp or fuzzy. That is, is
there a gradual or sharp transition where hydrogen goes from being mostly
ionized to mostly neutral?\\

\n The mean free path for photons traveling in an \HI region is simply
\begin{equation}
  \label{eq:strom:2} \ell \approx
  \frac{1}{n_{\mathrm{H^0}}\sigma_{\mathrm{\nu_0}}} \approx 0.05\
  \mathrm{\mathrm{pc}}\,\left(\frac{n_{\mathrm{H^0}}}{1\
  \mathrm{cm^{-3}}}\right)^{-1}
\end{equation}
This, compared to the sizes of \HII regions, is quite small, so we conclude the
transition to neutral hydrogen is rather sharp.\\

\n To actually solve for the ionization structure of the nebula, we need to use
the equation of radiative transfer:
\begin{equation}
  \label{eq:strom:3} \frac{dI_\nu}{ds} = -\sigma n_{\mathrm{H^0}}I_\nu + j_\nu
\end{equation}
where $J_\nu = 4\pi I_\nu$, and really the radiation field has two components,
that from the central star and the diffuse radiation field that arises from
recombinations in the [diffuse] nebula: $J_\nu = J_{\nu, *} + J_{\nu,
\mathrm{D}}$. First we'll analyze the field from the star:
\begin{equation}
  \label{eq:strom:4} 4\pi J_{\nu, *} = \pi B_{\nu, *}
   R_*\left(\frac{R_*}{R}\right)^2 e^{-\tau_\nu}
\end{equation}
Where the optical depth is
\begin{equation}
  \label{eq:strom:5} \tau_\nu = \int_0^R n_{\mathrm{H^0}}\sigma_{\mathrm{ph}}
  \,dR
\end{equation}
Now for the diffuse term:
\begin{equation}
  \label{eq:strom:6} 4\pi \int_{\nu_0}^\infty \frac{j_\nu}{h\nu} =
  n_{\mathrm{H^+}}n_e \alpha_1 = 4\pi \int_{\nu_0}^\infty n_{\mathrm{H}^0}
  \sigma \frac{J_{\nu, \mathrm{D}}}{h\nu}d\nu
\end{equation}
where $\alpha_1$ is the recombination coefficent that only counts
recombinations directly to the ground state, so
\begin{equation}
  \label{eq:strom:7} \alpha_1 \ll \alpha_{\mathrm{A}} = \sum_n\sum_L \alpha_{n,
  L}
\end{equation}
So the diffuse field is typically much weaker than the stellar radiation field,
which allows us to solve for the diffuse field iteratively. Again using our
``on the spot'' approxiamtion (photons from recombinations to $n=1$ are
immediately re-absorbed), the duffuse radiation field is
\begin{equation}
  \label{eq:strom:8} J_{\nu, \mathrm{D}} = \frac{j_\nu}{n_{\mathrm{H^0}}\sigma}
\end{equation}
and so at ionization equalibrium,
\begin{equation}
  \label{eq:strom:9} 4\pi \int \frac{J_{\nu, *}}{h\nu}\sigma n_{\mathrm{H^0}}\,
  d\nu + 4\pi \int\frac{J_{\nu,\mathrm{D}}}{h\nu}\sigma n_{\mathrm{H^0}}\,d\nu
  = n_{\mathrm{H^+}}n_e \alpha_{\mathrm{A}}(\mathrm{H},T)
\end{equation}
Simplifying with our approximations, this becomes
\begin{equation}
  \label{eq:strom:10} 4\pi \int \frac{J_{\nu, *}}{h\nu} \sigma n_{\mathrm{H^0}}
  \,d\nu = n_{\mathrm{H^+}}n_e\alpha_{\mathrm{B}}(\mathrm{H},T)
\end{equation}
We can now integrate outward from the star to find the ionization fraction,
defined by
\begin{equation}
  \label{eq:strom:11} \chi(\mathrm{H^+}) \equiv \frac{n(\mathrm{H^+})}{
  n(\mathrm{H^+}) + n(\mathrm{H}^0)}
\end{equation}
Converting \eqref{eq:strom:10} to variables that know about the star's surface
flux, it becomes
\begin{equation}
  \label{eq:strom:12} \boxed{\frac{\pi R_*^2}{R^2} \int
  \frac{F_{\nu,*}(R_*)e^{-\tau_\nu}}{h\nu} \sigma n_{\mathrm{H^0}}\,d\nu =
  n_{\mathrm{H^+}}n_e\alpha_{\mathrm{B}}(\mathrm{H},T)}
\end{equation}
where the optical depth is, scaled to the optical depth at line edge,
\begin{equation}
  \label{eq:strom:13} \tau = \frac{\sigma(\nu)}{\sigma(\nu_0)}\tau_0(r)
\end{equation}
% subsubsection ionization_equilibrium (end)
\subsubsection{The Str\"omgren Radius} % (fold)
\label{ssub:the_str"omgren_radius}
If we integrate \eqref{eq:strom:12} outwards and solve for the ionization
structure, we get
\begin{align}
  \label{eq:strom:14} \int_0^{R_{\mathrm{S}}} R^2_*\int n_{\mathrm{H^0}}
  \frac{\sigma}{h\nu} \frac{B_{\nu,*}(R_*)}{R^2}e^{-\tau_\nu} 4\pi R^2\,d\nu\,
  dR &= \int_0^{R_{\mathrm{S}}} n_e n_{\mathrm{H^+}}
  \alpha_{\mathrm{B}}(\mathrm{H}, T) 4\pi R^2\,dR\\
  \label{eq:strom:15} \int_{\nu_0}^\infty \pi R_*^2 \frac{B_{\nu,*}(R_*)}{h\nu}
  \int_0^1 e^{-\tau_\nu}d\tau\,d\nu &= \frac{4\pi}{3} R_{\mathrm{S}}^3 n_e
  n_{\mathrm{H^+}}\alpha_{\mathrm{B}}\\
  \label{eq:strom:16} 4\pi R_*^2 \int_{\nu_0}^\infty \frac{\pi B_{\nu, *}(R_*)}
  {h\nu} d\nu &= \frac{4\pi}{3} n_e n_{\mathrm{H^+}}R_{\mathrm{S}}^3
  \alpha_{\mathrm{B}}(\mathrm{H^0},T)
\end{align}
Where we've implicitly used the definition of the \textbf{Str\"omgren Radius} 
as the radius where the optical depth is unity. Physically, we can interpret the
left side of \eqref{eq:strom:16} as the total number of ionizations happening
as a result of the stellar radiation field. The right side of
\eqref{eq:strom:16} is just the total number of recombinations occuring within
the volume of the Str\"omgren sphere.\\

\n Defining the specific luminosity of the star simply as
\begin{equation}
  \label{eq:strom:17} L_{\nu, *} = 4\pi R_*^2 \pi B_{\nu, *}
\end{equation}
We can define the total number of ionizations more cleanly as 
\begin{equation}
  \label{eq:strom:18} Q = N_{\mathrm{ionizations}}(\mathrm{H^0}) \equiv 
  \int_{\nu_0}^\infty \frac{L_\nu}{h\nu}\,d\nu
\end{equation}
Or, in terms of the Str\"omgren radius,
\begin{equation}
  \label{eq:strom:19} \boxed{Q = \frac{4\pi}{3}R_{\mathrm{S}}^3
  \alpha_{\mathrm{B}}(\mathrm{H^0},T)n_en_{\mathrm{H^+}}}
\end{equation}
For a typical density ISM at $T\sim 10^4\ \mathrm{K}$, $Q$ is only dependent
on the stellar spectrum. Accordingly, we know that $Q$'s for $O$ stars are
significantly bigger than that for, say, $B$ stars. The same can be said about
the corrresponding Str\"omgren radii.\\

\n This concept can be extended to galactic scales by summing up the
contribution from all massive stars. For a galaxy with a star-forming rate of 1
$M_\odot\ \mathrm{yr}^{-1}$, the galactic $Q$ value is $1.08\times 10^{53}\
\mathrm{s^{-1}}$. We can then calculate the corresponding Str\"ombren radii for
galaxies.\\

\n As a more concrete example, consider a blackbody spectrum for our ionizing
source. Skipping the integrals, the Str\"omgren radius is simply
\begin{equation}
  \label{eq:strom:20} 78.2\ \mathrm{pc}\ \left(\frac{R_*}{R_\odot}\right)^{2/3}
  \left(\frac{n_{\mathrm{H^0}}}{1\ \mathrm{cm^{-3}}}\right)^{2/3}
  \frac{F_1(x_1)}{x_1}
\end{equation}
Where
\begin{equation}
  \label{eq:strom:21} x \equiv \frac{h\nu}{kT} \qquad \textrm{and} \qquad
  F_1(x_1) = \int_{x_1}^\infty \frac{x^2\,dx}{e^x-1}
\end{equation}
where $x_1$ is $x(\nu_0)$.
% subsubsection the_str"omgren_radius (end)
\subsubsection{Effects of Helium} % (fold)
\label{ssub:effects_of_helium}
So far we have only considered pure hydrogen, but really we must \emph{at least}
consider the impact of helium. If we only care about hte region of ionized
hydrogen, we'll need to start caring about which photons ionize helium and which
ionize hydrogen. We can still use the ``on the spot'' approximation for
recombinations to the ground state of helium. However, even recombinations to
some \emph{excited states} will also be immediately absorbed. The physics is
essentially the same, but the accounting is slightly more convoluted.
% subsubsection effects_of_helium (end)
% subsection stromgren_spheres (end)
% section h,ii_regions (end)
\end{document}

